---
title: "Estimating Generalized Linear Models for Continuous Data with rstanarm"
author: "Jonah Gabry and Ben Goodrich"
date: "`r Sys.Date()`"
output:
  html_vignette:
    toc: yes
---
<!--
%\VignetteEngine{knitr::rmarkdown}
%\VignetteIndexEntry{stan_glm: GLMs for Continuous Data}
-->

```{r, child="children/SETTINGS-knitr.txt"}
```
```{r, child="children/SETTINGS-gg.txt"}
```

# Introduction

This vignette explains how to estimate linear and generalized linear models 
(GLMs) for continuous response variables using the `stan_glm` function in the
__rstanarm__ package. For GLMs for discrete outcomes see the vignettes for
[binary/binomial](binomial.html) and [count](count.html) outcomes.

```{r, child="children/four_steps.txt"}
```

This vignette primarily focuses on Steps 1 and 2 when the likelihood is the
product of conditionally independent continuous distributions. Steps 3 and 4 are
covered in more depth by the vignette entitled ["How to Use the __rstanarm__
Package"](rstanarm.html), although this vignette does also give a few examples of model checking
and generating predictions.

# Likelihood

In the simplest case a GLM for a continuous outcome is simply a linear model 
and the likelihood for one observation is a conditionally normal PDF
$$\frac{1}{\sigma \sqrt{2 \pi}} e^{-\frac{1}{2} 
\left(\frac{y - \mu}{\sigma}\right)^2},$$
where $\mu = \alpha + \mathbf{x}^\top \boldsymbol{\beta}$ is a linear predictor
and $\sigma$ is the standard deviation of the error in predicting
the outcome, $y$.

More generally, a linear predictor $\eta = \alpha + \mathbf{x}^\top 
\boldsymbol{\beta}$ can be related to the conditional mean of the outcome via a 
link function $g$ that serves as a map between the range of values on which the 
outcome is defined and the space on which the linear predictor is defined. For 
the linear model described above no transformation is needed and so the link 
function is taken to be the identity function. However, there are cases in which
a link function is used for Gaussian models; the log link, for example, can be
used to log transform the (conditional) expected value of the outcome when it is
constrained to be positive.

Like the `glm` function, the `stan_glm` function uses R's family objects. The 
family objects for continuous outcomes compatible with `stan_glm` are the 
`gaussian`, `Gamma`, and `inverse.gaussian` distributions. All of the link 
functions provided by these family objects are also compatible with `stan_glm`. 
For example, for a Gamma GLM, where we assume that observations are
conditionally independent Gamma random variables, common link functions are the
log and inverse links.

Regardless of the distribution and link function, the likelihood for the entire
sample is the product of the likelihood contributions of the individual
observations.

# Priors

```{r, child="children/stan_glm_priors.txt"}
```

# Posterior

With independent prior distributions, the joint posterior distribution for
$\alpha$ and $\boldsymbol{\beta}$ is proportional to the product of the priors 
and the $N$ likelihood contributions:

$$f\left(\boldsymbol{\beta} | \mathbf{y},\mathbf{X}\right) \propto
  f\left(\alpha\right) \times \prod_{k=1}^K f\left(\beta_k\right) \times
  \prod_{i=1}^N {f(y_i|\eta_i)},$$
  
where $\mathbf{X}$ is the matrix of predictors and $\eta$ the linear predictor. 
This is the posterior distribution that `stan_glm` will draw from when using MCMC.
  
# Linear Regression Example

The `stan_lm` function, which has its own [vignette](lm.html), fits regularized linear
models using a novel means of specifying priors for the regression coefficients.
Here we focus using the `stan_glm` function, which can be used to estimate 
linear models with independent priors on the regression coefficients.

To illustrate the usage of `stan_glm` and some of the post-processing functions
in the __rstanarm__ package we'll use a simple example from Chapter 3 of [Gelman
and Hill (2007)](http://www.stat.columbia.edu/~gelman/arm/):

> We shall fit a series of regressions predicting cognitive test scores of
three- and four-year-old children given characteristics of their mothers, using
data from a survey of adult American women and their children (a subsample from
the National Longitudinal Survey of Youth).

Using two predictors -- a binary indicator for whether the mother has a
high-school degree (`mom_hs`) and the mother's score on an IQ test (`mom_iq`) --
we will fit four contending models. The first two models will each use just one
of the predictors, the third will use both, and the fourth will also include a
term for the interaction of the two predictors.

For these models we'll use the default weakly informative priors for `stan_glm`,
which are currently set to `normal(0,10)` for the intercept and `normal(0,5)`
for the other regression coefficients. For an overview of the many other
available prior distributions see `help("prior", package = "rstanarm")`.

```{r, continuous-kidiq-mcmc,results="hide"}
library(rstanarm)
data(kidiq)
post1 <- stan_glm(kid_score ~ mom_hs, data = kidiq, 
                  family = gaussian(link = "identity"), 
                  seed = 12345)
post2 <- update(post1, formula = . ~ mom_iq)
post3 <- update(post1, formula = . ~ mom_hs + mom_iq)
(post4 <- update(post1, formula = . ~ mom_hs * mom_iq))
```

```{r, continuous-kidiq-print, echo=FALSE}
print(post4)
```

Following Gelman and Hill's example, we make some plots overlaying the
estimated regression lines on the data.

```{r, continuous-kidiq-plot1a}
base <- ggplot(kidiq, aes(x = mom_hs, y = kid_score)) + 
  geom_point(size = 1, position = position_jitter(height = 0.05, width = 0.1)) + 
  scale_x_continuous(breaks = c(0,1), labels = c("No HS", "HS"))
  
base + geom_abline(intercept = coef(post1)[1], slope = coef(post1)[2], 
                   color = "skyblue4", size = 1)
```

There several ways we could add the uncertainty in our estimates to the plot. 
One way is to also plot the estimated regression line at each draw from the 
posterior distribution. To do this we can extract the posterior draws from the 
fitted model object using the `as.matrix` or `as.data.frame` methods:

```{r, continuous-kidiq-plot1b}
draws <- as.data.frame(post1)
colnames(draws)[1:2] <- c("a", "b")

base + 
  geom_abline(data = draws, aes(intercept = a, slope = b), 
              color = "skyblue", size = 0.2, alpha = 0.25) + 
  geom_abline(intercept = coef(post1)[1], slope = coef(post1)[2], 
              color = "skyblue4", size = 1)
```

For the second model we can make the same plot but the x-axis will show 
the continuous predictor `mom_iq`:

```{r, continuous-kidiq-plot2}
draws <- as.data.frame(as.matrix(post2))
colnames(draws)[1:2] <- c("a", "b")
ggplot(kidiq, aes(x = mom_iq, y = kid_score)) + 
  geom_point(size = 1) +
  geom_abline(data = draws, aes(intercept = a, slope = b), 
              color = "skyblue", size = 0.2, alpha = 0.25) + 
  geom_abline(intercept = coef(post2)[1], slope = coef(post2)[2], 
              color = "skyblue4", size = 1)
```

For the third and fourth models, each of which uses both predictors, we can plot
the continuous `mom_iq` on the x-axis and use color to indicate which 
points correspond to the different subpopulations defined by `mom_hs`. We also
now plot two regression lines, one for each subpopulation:

```{r, continuous-kidiq-plot3}
reg0 <- function(x, ests) cbind(1, 0, x) %*% ests 
reg1 <- function(x, ests) cbind(1, 1, x) %*% ests

args <- list(ests = coef(post3))
kidiq$clr <- factor(kidiq$mom_hs, labels = c("No HS", "HS"))
lgnd <- guide_legend(title = NULL)
base2 <- ggplot(kidiq, aes(x = mom_iq, fill = relevel(clr, ref = "HS"))) + 
  geom_point(aes(y = kid_score), shape = 21, stroke = .2, size = 1) + 
  guides(color = lgnd, fill = lgnd) + 
  theme(legend.position = "right")
base2 + 
  stat_function(fun = reg0, args = args, aes(color = "No HS"), size = 1.5) +
  stat_function(fun = reg1, args = args, aes(color = "HS"), size = 1.5)
```

```{r, continuous-kidiq-plot4}
reg0 <- function(x, ests) cbind(1, 0, x, 0 * x) %*% ests 
reg1 <- function(x, ests) cbind(1, 1, x, 1 * x) %*% ests
args <- list(ests = coef(post4))
base2 +
  stat_function(fun = reg0, args = args, aes(color = "No HS"), size = 1.5) + 
  stat_function(fun = reg1, args = args, aes(color = "HS"), size = 1.5)
```


## Model comparison 

One way we can compare the four contending models is to use an approximation to
Leave-One-Out (LOO) cross-validation, which is implemented by the `loo` function
in the __loo__ package:

```{r, continuous-kidiq-loo}
# Compare them with loo
loo1 <- loo(post1, cores = 2)
loo2 <- loo(post2, cores = 2)
loo3 <- loo(post3, cores = 2)
loo4 <- loo(post4, cores = 2)
(comp <- loo_compare(loo1, loo2, loo3, loo4))
```

In this case the fourth model is preferred as it has the highest 
expected log predicted density (`elpd_loo`) or, equivalently, the lowest
value of the LOO Information Criterion (`looic`). The fourth model
is preferred by a lot over the first model

```{r, continuous-kidiq-loo-2}
loo_compare(loo1, loo4)
```

because the difference in `elpd` is so much larger than the standard error. 
However, the preference of the fourth model over the others isn't as strong:

```{r, continuous-kidiq-loo-3}
loo_compare(loo3, loo4)
loo_compare(loo2, loo4)
```

## The posterior predictive distribution 

The posterior predictive distribution is the distribution of the outcome implied
by the model after using the observed data to update our beliefs about the 
unknown parameters. When simulating observations from the posterior predictive 
distribution we use the notation $y^{\rm rep}$ (for _replicate_) when we use the
same observations of $X$ that were used to estimate the model parameters. When
$X$ contains new observations we use the notation $\tilde{y}$ to refer to the 
posterior predictive simulations.

Simulating data from the posterior predictive distribution using the observed
predictors is useful for checking the fit of the model. Drawing from the
posterior predictive distribution at interesting values of the predictors also
lets us visualize how a manipulation of a predictor affects (a function of) the
outcome(s).

### Graphical posterior predictive checks

The `pp_check` function generates a variety of plots comparing the observed 
outcome $y$ to simulated datasets $y^{\rm rep}$ from the posterior predictive 
distribution using the same observations of the predictors $X$ as we used to fit
the model. He we show a few of the possible displays. The documentation at
`help("pp_check.stanreg", package = "rstanarm")` has details on all of the
`pp_check` options.

First we'll look at a plot directly comparing the distributions of $y$ and
$y^{\rm rep}$. The following call to `pp_check` will create a plot juxtaposing
the histogram of $y$ and histograms of five $y^{\rm rep}$ datasets:

```{r, continuous-kidiq-pp_check1}
pp_check(post4, plotfun = "hist", nreps = 5)
```

The idea is that if the model is a good fit to the data we should be able to 
generate data $y^{\rm rep}$ from the posterior predictive distribution that 
looks a lot like the observed data $y$. That is, given $y$, the $y^{\rm
rep}$ we generate should be plausible.

Another useful plot we can make using `pp_check` shows the distribution of a test
quantity $T(y^{\rm rep})$ compared to $T(y)$, the value of the quantity in the 
observed data. When the argument `plotfun = "stat"` is specified, `pp_check` will 
simulate $S$ datasets $y_1^{\rm rep}, \dots, y_S^{\rm rep}$, each containing $N$
observations. Here $S$ is the size of the posterior sample (the number of MCMC 
draws from the posterior distribution of the model parameters) and $N$ is the 
length of $y$. We can then check if $T(y)$ is consistent with the distribution 
of $\left(T(y_1^{\rm yep}), \dots, T(y_S^{\rm yep})\right)$. In the plot below
we see that the mean of the observations is plausible when compared to the
distribution of the means of the $S$ $y^{\rm rep}$ datasets:

```{r, continuous-kidiq-pp_check2}
pp_check(post4, plotfun = "stat", stat = "mean")
```

Using `plotfun="stat_2d"` we can also specify two test quantities and look at a
scatterplot:

```{r, continuous-kidiq-pp_check3}
pp_check(post4, plotfun = "stat_2d", stat = c("mean", "sd"))
```

### Generating predictions

The `posterior_predict` function is used to generate replicated data $y^{\rm
rep}$ or predictions for future observations $\tilde{y}$. Here we show how to
use `posterior_predict` to generate predictions of the outcome `kid_score` for a
range of different values of `mom_iq` and for both subpopulations defined by
`mom_hs`.


```{r, continuous-kidiq-posterior_predict}
IQ_SEQ <- seq(from = 75, to = 135, by = 5)
y_nohs <- posterior_predict(post4, newdata = data.frame(mom_hs = 0, mom_iq = IQ_SEQ))
y_hs <- posterior_predict(post4, newdata = data.frame(mom_hs = 1, mom_iq = IQ_SEQ))
dim(y_hs)
```

We now have two matrices, `y_nohs` and `y_hs`. Each matrix has 
as many columns as there are values of `IQ_SEQ` and as many rows as
the size of the posterior sample. One way to show the predictors is to plot the predictions 
for the two groups of kids side by side:

```{r, continuous-kidiq-plot-predict, fig.width=7}
par(mfrow = c(1:2), mar = c(5,4,2,1))
boxplot(y_hs, axes = FALSE, outline = FALSE, ylim = c(10,170),
        xlab = "Mom IQ", ylab = "Predicted Kid IQ", main = "Mom HS")
axis(1, at = 1:ncol(y_hs), labels = IQ_SEQ, las = 3)
axis(2, las = 1)
boxplot(y_nohs, outline = FALSE, col = "red", axes = FALSE, ylim = c(10,170),
        xlab = "Mom IQ", ylab = NULL, main = "Mom No HS")
axis(1, at = 1:ncol(y_hs), labels = IQ_SEQ, las = 3)
```

```{r, continuous-kidiq-validation, eval=FALSE, include=FALSE}
# # External Validation
# source(paste0(ROOT, "ARM/Ch.3/kids_before1987.data.R"), 
#        local = kidiq, verbose = FALSE)
# source(paste0(ROOT, "ARM/Ch.3/kids_after1987.data.R"), 
#        local = kidiq, verbose = FALSE)
# post5 <- stan_lm(ppvt ~ hs + afqt, data = kidiq,
#                  prior = R2(location = 0.25, what = "mean"), seed = SEED)
# y_ev <- posterior_predict(post5, newdata = 
#                           data.frame(hs = kidiq$hs_ev, afqt = kidiq$afqt_ev))
# par(mfrow = c(1,1))
# hist(-sweep(y_ev, 2, STATS = kidiq$ppvt_ev, FUN = "-"), prob = TRUE,
#      xlab = "Predictive Errors in ppvt", main = "", las = 2)
```

# Gamma Regression Example

Gamma regression is often used when the response variable is continuous and positive, 
and the _coefficient of variation_ (rather than the variance) is constant.

We'll use one of the standard examples of Gamma regression, 
which is taken from McCullagh & Nelder (1989). This example is also given in the
documentation for R's `glm` function. The outcome of interest is the clotting
time of blood (in seconds) for "normal plasma diluted to nine different
percentage concentrations with prothrombin-free plasma; clotting was induced by
two lots of thromboplastin" (p. 300).

The help page for R's `glm` function presents the example as follows:

```{r, continuous-clotting-mle, results='hide'}
clotting <- data.frame(
    u = c(5,10,15,20,30,40,60,80,100),
    lot1 = c(118,58,42,35,27,25,21,19,18),
    lot2 = c(69,35,26,21,18,16,13,12,12))
summary(glm(lot1 ~ log(u), data = clotting, family = Gamma))
summary(glm(lot2 ~ log(u), data = clotting, family = Gamma))
```

To fit the analogous Bayesian models we can simply substitute `stan_glm` for
`glm` above. However, instead of fitting separate models we can also reshape the
data slightly and fit a model interacting lot with plasma concentration:

```{r, continuous-clotting-mcmc, results="hide"}
clotting2 <- with(clotting, data.frame(
  log_plasma = rep(log(u), 2),
  clot_time = c(lot1, lot2),
  lot_id = factor(rep(c(1,2), each = length(u)))
))

fit <- stan_glm(clot_time ~ log_plasma * lot_id, data = clotting2, family = Gamma, 
                prior_intercept = normal(0,1), prior = normal(0,1),
                seed = 12345)
```

```{r}
print(fit, digits = 3)
```

In the output above, the estimate reported for `shape` is for the shape 
parameter of the Gamma distribution. The _reciprocal_ of the shape parameter can
be interpreted similarly to what `summary.glm` refers to as the dispersion
parameter.



# References
Gelman, A. and Hill, J. (2007). _Data Analysis Using Regression and
Multilevel/Hierarchical Models._ Cambridge University Press, Cambridge, UK.

McCullagh, P. and Nelder, J. A. (1989). _Generalized Linear Models._ Chapman and
Hall/CRC Press, New York.
