---
title: "Estimating Generalized Linear Models for Count Data with rstanarm"
author: "Jonah Gabry and Ben Goodrich"
date: "`r Sys.Date()`"
output:
  html_vignette:
    toc: yes
---
<!--
%\VignetteEngine{knitr::rmarkdown}
%\VignetteIndexEntry{stan_glm: GLMs for Count Data}
-->
```{r, child="children/SETTINGS-knitr.txt"}
```
```{r, child="children/SETTINGS-gg.txt"}
```

# Introduction

This vignette explains how to estimate generalized linear models (GLMs) for 
count data using the `stan_glm` function in the __rstanarm__ package.

```{r, child="children/four_steps.txt"}
```

Steps 3 and 4 are covered in more depth by the vignette entitled ["How to Use the
__rstanarm__ Package"](rstanarm.html). This vignette focuses on Step 1 for Poisson and negative
binomial regression models using the `stan_glm` function.

# Likelihood

If the outcome for a single observation $y$ is assumed to follow a Poisson
distribution, the likelihood for one observation can be written as a
conditionally Poisson PMF

$$\tfrac{1}{y!} \lambda^y e^{-\lambda},$$

where $\lambda = E(y | \mathbf{x}) = g^{-1}(\eta)$ and $\eta = \alpha +
\mathbf{x}^\top \boldsymbol{\beta}$ is a linear predictor. For the Poisson 
distribution it is also true that $\lambda = Var(y | \mathbf{x})$, i.e. the 
mean and variance are both $\lambda$. Later in this vignette we also show how
to estimate a negative binomial regression, which relaxes this assumption of 
equal conditional mean and variance of $y$. 

Because the rate parameter $\lambda$ must be positive, for a Poisson GLM the
_link_ function $g$ maps between the positive real numbers $\mathbb{R}^+$ (the
support of $\lambda$) and the set of all real numbers $\mathbb{R}$. When applied
to a linear predictor $\eta$ with values in $\mathbb{R}$, the inverse link
function $g^{-1}(\eta)$ therefore returns a positive real number.

Although other link functions are possible, the canonical link function for a 
Poisson GLM is the log link $g(x) = \ln{(x)}$. With the log link, the inverse
link function is simply the exponential function and the likelihood for a single
observation becomes

$$\frac{g^{-1}(\eta)^y}{y!} e^{-g^{-1}(\eta)} = 
\frac{e^{\eta y}}{y!} e^{-e^\eta}.$$

# Priors

```{r, child="children/stan_glm_priors.txt"}
```

# Posterior

With independent prior distributions, the joint posterior distribution for 
$\alpha$ and $\boldsymbol{\beta}$ in the Poisson model is proportional to the
product of the priors and the $N$ likelihood contributions:

$$f\left(\alpha,\boldsymbol{\beta} | \mathbf{y},\mathbf{X}\right) \propto
  f\left(\alpha\right) \times \prod_{k=1}^K f\left(\beta_k\right) \times
  \prod_{i=1}^N {
  \frac{g^{-1}(\eta_i)^{y_i}}{y_i!} e^{-g^{-1}(\eta_i)}}.$$
  
This is posterior distribution that `stan_glm` will draw from when using MCMC.

# Poisson and Negative Binomial Regression Example

This example comes from Chapter 8.3 of 
[Gelman and Hill (2007)](http://www.stat.columbia.edu/~gelman/arm/). 

We want to make inferences about the efficacy of a certain pest 
management system at reducing the number of roaches in urban apartments. Here is
how Gelman and Hill describe the experiment (pg. 161):

> [...] the treatment and control were applied to 160 and 104 apartments, 
respectively, and the outcome measurement $y_i$ in each apartment $i$ was the 
number of roaches caught in a set of traps. Different apartments had traps for 
different numbers of days [...] 

In addition to an intercept, the regression predictors for the model are the 
pre-treatment number of roaches `roach1`, the treatment indicator 
`treatment`, and a variable indicating whether the apartment is in a building
restricted to elderly residents `senior`. Because the number of days for which
the roach traps were used is not the same for all apartments in the sample, we
include it as an exposure, which slightly changes the model described in
the __Likelihood__ section above in that the rate parameter $\lambda_i =
exp(\eta_i)$ is multiplied by the exposure $u_i$ giving us 
$y_i \sim Poisson(u_i \lambda_i)$. This is equivalent to adding $\ln{(u_i)}$ 
to the linear predictor $\eta_i$ and it can be specified using the `offset` 
argument to `stan_glm`.

```{r, count-roaches-mcmc, results="hide"}
library(rstanarm)
data(roaches)

# Rescale
roaches$roach1 <- roaches$roach1 / 100
# Estimate original model
glm1 <- glm(y ~ roach1 + treatment + senior, offset = log(exposure2), 
            data = roaches, family = poisson)
# Estimate Bayesian version with stan_glm
stan_glm1 <- stan_glm(y ~ roach1 + treatment + senior, offset = log(exposure2),
                      data = roaches, family = poisson, 
                      prior = normal(0, 2.5, autoscale=FALSE), 
                      prior_intercept = normal(0, 5, autoscale=FALSE),
                      seed = 12345)
```

The `formula`, `data`, `family`, and `offset` arguments to `stan_glm` can be 
specified in exactly the same way as for `glm`. The `poisson` family function 
defaults to using the log link, but to write code readable to someone not 
familiar with the defaults we should be explicit and use 
`family = poisson(link = "log")`.  

We've also specified some optional arguments. The `chains` argument controls how
many Markov chains are executed, the `cores` argument controls the number of
cores utilized by the computer when fitting the model. We also provided a seed
so that we have the option to deterministically reproduce these results at any
time. The `stan_glm` function has many other optional arguments that allow for
more user control over the way estimation is performed. The documentation for
`stan_glm` has more information about these controls as well as other topics
related to GLM estimation.

Here are the point estimates and uncertainties from the `glm` fit and `stan_glm`
fit, which we see are nearly identical:

```{r, count-roaches-comparison}
round(rbind(glm = coef(glm1), stan_glm = coef(stan_glm1)), digits = 2)
round(rbind(glm = summary(glm1)$coefficients[, "Std. Error"], 
            stan_glm = se(stan_glm1)), digits = 3)
```

(Note: the dataset we have is slightly different from the one used in Gelman and
Hill (2007), which leads to slightly different parameter estimates than those
shown in the book even when copying the `glm` call verbatim. Also, we have
rescaled the `roach1` predictor. For the purposes of this example, the actual 
estimates are less important than the process.)

Gelman and Hill next show how to compare the observed data to replicated 
datasets from the model to check the quality of the fit. Here we don't show the
original code used by Gelman and Hill because it's many lines, requiring several
loops and some care to get the matrix multiplications right (see pg. 161-162).
On the other hand, the __rstanarm__ package makes this easy. We can generate
replicated datasets with a single line of code using the `posterior_predict`
function:

```{r, count-roaches-posterior_predict}
yrep <- posterior_predict(stan_glm1)
```

By default `posterior_predict` will generate a dataset for each set of 
parameter draws from the posterior distribution. That is, `yrep` will be an 
$S \times N$ matrix, where $S$ is the size of the posterior sample and $N$
is the number of data points. Each row of `yrep` represents a full dataset
generated from the posterior predictive distribution. For more about the 
importance of the `posterior_predict` function, see the 
["How to Use the __rstanarm__ Package"](rstanarm.html) vignette.

Gelman and Hill take the simulated datasets and for each of them compute the 
proportion of zeros and compare to the observed proportion in the original 
data. We can do this easily using the `pp_check` function, which generates 
graphical comparisons of the data `y` and replicated datasets `yrep`.

```{r, count-roaches-plot-pp_check1}
prop_zero <- function(y) mean(y == 0)
(prop_zero_test1 <- pp_check(stan_glm1, plotfun = "stat", stat = "prop_zero", binwidth = .005))
```

The value of the test statistic (in this case the proportion of zeros) computed
from the sample `y` is the dark blue vertical line. More than 30% of these
observations are zeros, whereas the replicated datasets all contain less than 1%
zeros (light blue histogram). This is a sign that we should consider a model
that more accurately accounts for the large proportion of zeros in the data.
Gelman and Hill show how we can do this using an overdispersed Poisson
regression. To illustrate the use of a different `stan_glm` model, here we will
instead try
[negative binomial](https://en.wikipedia.org/wiki/Negative_binomial_distribution) 
regression, which is also used for overdispersed or zero-inflated count data.
The negative binomial distribution allows the (conditional) mean and variance of
$y$ to differ unlike the Poisson distribution. To fit the negative binomial
model can either use the `stan_glm.nb` function or, equivalently, change the
`family` we specify in the call to `stan_glm` to `neg_binomial_2` instead of
`poisson`. To do the latter we can just use `update`:

```{r, count-roaches-negbin, results="hide"}
stan_glm2 <- update(stan_glm1, family = neg_binomial_2) 
```

We now use `pp_check` again, this time to check the proportion of zeros in the
replicated datasets under the negative binomial model:

```{r, count-roaches-plot-pp_check2, fig.width=7, out.width="80%"}
prop_zero_test2 <- pp_check(stan_glm2, plotfun = "stat", stat = "prop_zero", 
                            binwidth = 0.01)
# Show graphs for Poisson and negative binomial side by side
bayesplot_grid(prop_zero_test1 + ggtitle("Poisson"), 
               prop_zero_test2 + ggtitle("Negative Binomial"), 
               grid_args = list(ncol = 2))
```

This is a much better fit, as the proportion of zeros in the data falls nicely 
near the center of the distribution of the proportion of zeros among the
replicated datasets. The observed proportion of zeros is quite plausible under
this model.

We could have also made these plots manually without using the `pp_check` function 
because we have the `yrep` datasets created by `posterior_predict`. The `pp_check`
function takes care of this for us, but `yrep` can be used directly to carry out 
other posterior predictive checks that aren't automated by `pp_check`. 

When we comparing the models using the __loo__ package we also see a clear
preference for the negative binomial model

```{r, count-roaches-loo}
loo1 <- loo(stan_glm1, cores = 2)
loo2 <- loo(stan_glm2, cores = 2)
loo_compare(loo1, loo2)
```

which is not surprising given the better fit we've already observed from the
posterior predictive checks.


# References

Gelman, A. and Hill, J. (2007). _Data Analysis Using Regression and
Multilevel/Hierarchical Models._ Cambridge University Press, Cambridge, UK.
