---
title: "Estimating Joint Models for Longitudinal and Time-to-Event Data with rstanarm"
author: "Sam Brilleman"
date: "`r Sys.Date()`"
output: 
  html_vignette:
    toc: yes
params:
  EVAL: !r identical(Sys.getenv("NOT_CRAN"), "true")
---
<!--
%\VignetteEngine{knitr::rmarkdown}
%\VignetteIndexEntry{stan_jm: Joint Models for Longitudinal and Time-to-Event Data}
-->
```{r, child="children/SETTINGS-knitr.txt"}
```
```{r, child="children/SETTINGS-gg.txt"}
```
```{r, child="children/SETTINGS-rstan.txt"}
```
```{r, child="children/SETTINGS-loo.txt"}
```

## Introduction

This vignette provides an introduction to the `stan_jm` modelling function in the __rstanarm__ package. The `stan_jm` function allows the user to estimate a shared parameter joint model for longitudinal and time-to-event data under a Bayesian framework. 

Within a shared parameter joint model there are several outcomes: one or more *longitudinal* outcomes, and one *time-to-event* outcome. Note that the longitudinal outcome(s) are sometimes referred to as longitudinal *biomarkers*, since they commonly represent a clinical biomarker which is repeatedly measured over time (for example blood pressure). In most situations, we will be interested in how the longitudinal biomarker, or some aspect of the trajectory of the longitudinal biomarker, is associated with the risk of the event. The joint modelling approach described below, and implemented in the `stan_jm` modelling function, allows us to estimate this association allowing for the fact that the observed longitudinal measurements are likely to be both error-prone and measured in discrete time, whilst underlying changes in the unobserved "true" value of the biomarker are both free of measurement error and occur in continuous time. 

Note that some aspects of the estimation are covered in other vignettes, such as the `stan_glmer` [vignette](glmer.html) which contains details on the prior distribution for covariance matrices for the group-specific terms, or the priors [vignette](priors.html) which contains details on the prior distributions available for regression coefficients.

## Model formulation

A shared parameter joint model consists of related submodels which are specified separately for each of the outcomes. These are therefore commonly referred to as the *longitudinal submodel(s)* and the *event submodel*. The longitudinal and event submodels are linked using shared individual-specific parameters, which can be parameterised in a number of ways. The specific model formulation used by the `stan_jm` modelling function is described below.

### Longitudinal submodel(s)

We assume $y_{ijm}(t) = y_{im}(t_{ij})$ corresponds to the observed value
of the $m^{th} (m = 1,...,M)$ biomarker for individual $i (i = 1,...,N)$ taken 
at time point $t_{ij}, j =  1,...,n_i$. We specify a (multivariate) 
generalised linear mixed model that assumes $y_{ijm}(t)$ follows a 
distribution in the exponential family with mean $\mu_{ijm}(t)$ and linear 
predictor 

$$
\eta_{ijm}(t) = g_m(\mu_{ijm}(t)) = 
  \boldsymbol{x}^T_{ijm}(t) \boldsymbol{\beta}_m +
  \boldsymbol{z}^T_{ijm}(t) \boldsymbol{b}_{im}
$$

where $\boldsymbol{x}^T_{ijm}(t)$ and $\boldsymbol{z}^T_{ijm}(t)$ are both
row-vectors of covariates (which likely include some function of time, for 
example a linear slope, cubic splines, or polynomial terms) with associated 
vectors of fixed and individual-specific parameters $\boldsymbol{\beta}_m$ and 
$\boldsymbol{b}_{im}$, respectively, and $g_m$ is some known link function. 

The distribution and link function are allowed to differ over the $M$ longitudinal
submodels. We assume that the dependence across the different longitudinal
submodel (i.e. the correlation between the different longitudinal biomarkers)
is captured through a shared multivariate normal distribution for the
individual-specific parameters; that is, we assume 

$$
\begin{pmatrix} \boldsymbol{b}_{i1} \\ \vdots \\ \boldsymbol{b}_{iM} \end{pmatrix} = 
  \boldsymbol{b}_i \sim
  \mathsf{Normal} \left( 0 , \boldsymbol{\Sigma} \right)
$$

for some unstructured variance-covariance matrix $\boldsymbol{\Sigma}$ 
(see the `stan_glmer` [vignette](glmer.html) for details of the prior 
distribution used for $\boldsymbol{\Sigma}$).

### Event submodel

We assume that we also observe an event time 
$T_i = \mathsf{min} \left( T^*_i , C_i \right)$ where $T^*_i$ denotes the 
so-called "true" event time for individual $i$ (potentially unobserved) and 
$C_i$ denotes the censoring time. We define an event indicator 
$d_i = I(T^*_i \leq C_i)$. We then model the hazard of the event using a 
parametric proportional hazards regression model of the form

$$
h_i(t) = h_0(t) \mathsf{exp} 
  \left(
    \boldsymbol{w}^T_i(t) \boldsymbol{\gamma} +
    \sum_{m=1}^M \sum_{q=1}^{Q_m} \alpha_{mq}
      f_{mq}(\boldsymbol{\beta}_m, \boldsymbol{b}_{im}; t) 
  \right) 
$$

where $h_i(t)$ is the hazard of the event for individual $i$ at time $t$, 
$h_0(t)$ is the baseline hazard at time $t$, $\boldsymbol{w}^T_i(t)$ is a 
row-vector of individual-specific covariates (possibly time-dependent) with 
an associated vector of regression coefficients $\boldsymbol{\gamma}$ (log
hazard ratios).

The longitudinal and event submodels are assumed to related via an "association 
structure" based on shared individual-specific parameters and captured via the
$\sum_{m=1}^M \sum_{q=1}^{Q_m} \alpha_{mq} f_{mq}(\boldsymbol{\beta}_m, \boldsymbol{b}_{im}; t)$ 
term in the linear predictor of the proportional hazards regression model. 
The $\alpha_{mq}$ are referred to as the "association parameters" since they quantify the strength of the association between the longitudinal and event processes, while the $f_{mq}(\boldsymbol{\beta}_m, \boldsymbol{b}_{im}; t)$ (for some functions $f_{mq}(.)$) can be referred to as the "association terms" and can be specified in a variety of ways which we describe in the next section.

Note that in the `stan_jm` modelling function the baseline hazard be specified as either a Weibull distribution (the default), a piecewise constant baseline hazard (sometimes referred to as piecewise exponential), or an approximation using B-splines (the choice can be made via the `basehaz` argument). In the case of the piecewise constant or B-splines baseline hazard, the user can control the flexibility by specifying the knots or degrees of freedom via the `basehaz_ops` argument.

### Association structures

The longitudinal submodel(s) and the event submodel are assumed to be related through shared group-specific parameters. Technically, a key assumption of the shared parameter joint model is that the longitudinal outcome(s) are independent of the event time conditional on the individual-specific parameters $\boldsymbol{b}_i$. As mentioned in the previous section, the shared parameters can be specified in a variety of ways and the specification we choose is commonly referred to as the *association structure* of the joint model. The `stan_jm` modelling function allows for the following association structures:

Current value (of the linear predictor or expected value)
$$
f_{mq}(\boldsymbol{\beta}_m, \boldsymbol{b}_{im}; t) = \eta_{ijm}(t) \\
f_{mq}(\boldsymbol{\beta}_m, \boldsymbol{b}_{im}; t) = \mu_{ijm}(t)
$$
Current slope (of the linear predictor or expected value)
$$
f_{mq}(\boldsymbol{\beta}_m, \boldsymbol{b}_{im}; t) = \frac{d\eta_{ijm}(t)}{dt} \\
f_{mq}(\boldsymbol{\beta}_m, \boldsymbol{b}_{im}; t) = \frac{d\mu_{ijm}(t)}{dt}
$$
Area under the curve (of the linear predictor or expected value)
$$
f_{mq}(\boldsymbol{\beta}_m, \boldsymbol{b}_{im}; t) = \int_0^t \eta_{ijm}(u) du \\
f_{mq}(\boldsymbol{\beta}_m, \boldsymbol{b}_{im}; t) = \int_0^t \mu_{ijm}(u) du
$$
Interactions between different biomarkers
$$
f_{mq}(\boldsymbol{\beta}_m, \boldsymbol{b}_{im}; t) = \eta_{ijm}(t) \eta_{ijm'}(t) 
  \text{ for some } m = m' \text{ or } m \neq m' \\
f_{mq}(\boldsymbol{\beta}_m, \boldsymbol{b}_{im}; t) = \eta_{ijm}(t) \mu_{ijm'}(t) 
  \text{ for some } m = m' \text{ or } m \neq m' \\
f_{mq}(\boldsymbol{\beta}_m, \boldsymbol{b}_{im}; t) = \mu_{ijm}(t) \mu_{ijm'}(t) 
  \text{ for some } m = m' \text{ or } m \neq m'
$$   
Interactions between the biomarker (or it's slope) and observed data
$$
f_{mq}(\boldsymbol{\beta}_m, \boldsymbol{b}_{im}; t) = \eta_{ijm}(t) c_{i}(t) 
  \text{  for some covariate value } c_{i}(t) \\
f_{mq}(\boldsymbol{\beta}_m, \boldsymbol{b}_{im}; t) = \mu_{ijm}(t) c_{i}(t) 
  \text{  for some covariate value } c_{i}(t) \\
f_{mq}(\boldsymbol{\beta}_m, \boldsymbol{b}_{im}; t) = \frac{d\eta_{ijm}(t)}{dt} c_{i}(t) 
  \text{  for some covariate value } c_{i}(t) \\
f_{mq}(\boldsymbol{\beta}_m, \boldsymbol{b}_{im}; t) = \frac{d\mu_{ijm}(t)}{dt} c_{i}(t) 
  \text{  for some covariate value } c_{i}(t)
$$

As well as using lagged values for any of the above. That is, replacing $t$ with
$t-u$ where $u$ is some lag time, such that the hazard of the event at time $t$ 
is assumed to be associated with some function of the longitudinal submodel
parameters at time $t-u$.

Lastly, we can specify some time-independent function of the random effects,
possibly including the fixed effect component. For example, 

$$
f_{mq}(\boldsymbol{\beta}_m, \boldsymbol{b}_{im}; t) = 
  \boldsymbol{b}_{im0}
$$
or 
$$
f_{mq}(\boldsymbol{\beta}_m, \boldsymbol{b}_{im}; t) = 
  \boldsymbol{\beta}_{m0} + \boldsymbol{b}_{im0}
$$
where $\boldsymbol{\beta}_{m0}$ is the population-level intercept for the $m^{th}$
longitudinal submodel and $\boldsymbol{b}_{im0}$ is the $i^{th}$ individual's 
random deviation from the population-level intercept for the $m^{th}$ longitudinal 
submodel.

Note that more than one association structure can be specified, however, not all possible combinations are allowed. Moreover, if you are fitting a multivariate joint model (i.e. more than one longitudinal outcome) then you can optionally choose to use a different association structure(s) for linking each longitudinal submodel to the event submodel. To do this you can pass a list of length $M$ to the `assoc` argument.

### Likelihood function

Not yet completed

## Example: PBC dataset

### The simplest univariate joint model

In this section we present an example using the Mayo Clinic's primary biliary cirrhosis (PBC) data. So that the examples run quickly, we use a small random subset of just 40 patients from the full data. For a description of the dataset you can type:

```{r datasets_help, eval = FALSE}
help("datasets", package = "rstanarm")
```

We first fit a simple univariate joint model, with a single normally distributed longitudinal marker, an association structure based on the current value of the linear predictor, and Weibull baseline hazard. To fit the model we use the main modelling function in the **rstanjm** package: `stan_jm`. When calling `stan_jm` we must, at a minimum, specify a formula object for each of the longitudinal and event submodels (through the arguments `formulaLong` and `formulaEvent`), the data frames which contain the variables for each of the the longitudinal and event submodels (through the arguments `dataLong` and `dataEvent`), and the name of the variable representing time in the longitudinal submodel (through the argument `time_var`).

The formula for the longitudinal submodel is specified using the **lme4** package formula style. That is `y ~ x + (random_effects | grouping_factor)`. In this example we specify that log serum bilirubin (`logBili`) follows a subject-specific linear trajectory. To do this we include a fixed intercept and fixed slope (`year`), as well as a random intecept and random slope for each subject `id` (`(year | id)`).

The formula for the event submodel is specified using the **survival** package formula style. That is, the outcome of the left of the `~` needs to be of the format `Surv(event_time, event_indicator)` for single row per individual data, or `Surv(start_time, stop_time, event_indicator)` for multiple row per individual data. The latter allows for exogenous time-varying covariates to be included in the event submodel. In this example we assume that the log hazard of death is linearly related to gender (`sex`) and an indicator of treatment with D-penicillamine (`trt`).

```{r univariate_fit, results = "hold", message = FALSE, warning = FALSE}
library(rstanarm)
f1 <- stan_jm(formulaLong = logBili ~ year + (year | id), 
              dataLong = pbcLong,
              formulaEvent = survival::Surv(futimeYears, death) ~ sex + trt, 
              dataEvent = pbcSurv,
              time_var = "year",
              chains = 1, refresh = 2000, seed = 12345)
```

The argument `refresh = 2000` was specified so that Stan didn't provide us with excessive progress updates whilst fitting the model. However, if you are fitting a model that will take several minutes or hours to fit, then you may wish to request progress updates quite regularly, for example setting `refresh = 20` for every 20 iterations (by default the refresh argument is set to 1/10th of the total number of iterations). 

The fitted model is returned as an object of the S3 class `stanjm`. We have a variety of methods and postestimation functions available for this class, including: `print`, `summary`, `plot`, `fixef`, `ranef`, `coef`, `VarCorr`, `posterior_interval`, `update`, and more. Here, we will examine the most basic output for the fitted joint model by typing `print(f1)`:

```{r print, echo = FALSE}
print(f1)
```

The output tells us that for each one unit increase in an individual's underlying level of log serum bilirubin, their estimated log hazard of death increases by 34% (equivalent to a 3.8-fold increase in the hazard of death). The mean absolute deviation (MAD) is provided as a more robust estimate of the standard deviation of the posterior distribution. In this case the MAD_SD for the association parameter is 0.237, indicating there is quite large uncertainty around the estimated association between log serum bilirubin and risk of death (recall this is a small dataset containing only 40 patients!).

If we wanted some slightly more detailed output for each of the model parameters, as well as further details regarding the model estimation (for example computation time, number of longitudinal observations, number of individuals, type of baseline hazard, etc) we can instead use the `summary` method:

```{r summary}
summary(f1)
```

The easiest way to extract the correlation matrix for the random effects (aside from viewing the `print` output) is to use the `VarCorr` function (modelled on the `VarCorr` function from the **lme4** package). If you wish to extract the variances and covariances (instead of the standard deviations and correlations) then you can type the following to return a data frame with all of the relevant information:

```{r VarCorr}
as.data.frame(VarCorr(f1))
```

### Current slope association structure

In the previous example we were fitting a shared parameter joint model which assumed that the log hazard of the event (in this case the log hazard of death) at time *t* was linearly related to the subject-specific expected value of the longitudinal marker (in this case the expected value of log serum bilirubin) also at time *t*. This is the default association structure, although it could be explicitly specified by setting the `assoc = "etavalue"` argument. 

However, let's suppose we believe that the log hazard of death is actually related to both the *current value* of log serum bilirubin and the current *rate of change* in log serum bilirubin. To estimate this joint model we need to indicate that we want to also include the subject-specific slope (at time *t*) from the longitudinal submodel as part of the association structure. We do this by setting the `assoc` argument equal to a character vector `c("etavalue", "etaslope")` which indicates our desired association structure:

```{r assoc_etaslope, eval = FALSE}
f2 <- stan_jm(formulaLong = logBili ~ year + (year | id), 
              dataLong = pbcLong,
              formulaEvent = survival::Surv(futimeYears, death) ~ sex + trt, 
              dataEvent = pbcSurv,
              assoc = c("etavalue", "etaslope"),
              time_var = "year", 
              chains = 1, refresh = 2000, seed = 12345)
``` 

In this example the subject-specific slope is actually constant across time *t* since we have a linear trajectory. Note however that we could still use the `"etaslope"` association structure even if we had a non-linear subject specific trajectory (for example modelled using cubic splines or polynomials).

If we instead thought that the log hazard of death was only related to how an individual's log serum bilirubin deviated from the average at baseline, we could form an association structure based only on the subject-specific random intercept term. To do this we would use the `"shared_b"` association structure. However, by default this would include all of the subject-specific random effects; in this example we have a random intercept and a random slope. To only include a subset of the random effects in the association structure we can specify the indices in parentheses as a suffix, as follows:

```{r assoc_randomint, eval = FALSE}
f3 <- stan_jm(formulaLong = logBili ~ year + (year | id), 
              dataLong = pbcLong,
              formulaEvent = survival::Surv(futimeYears, death) ~ sex + trt, 
              dataEvent = pbcSurv,
			        assoc = "shared_b(1)",
			        time_var = "year", 
              chains = 1, refresh = 2000, seed = 12345))
```


